**The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models**

### 2021-03-08 Manuscripts/Manuscript_AM_v3.Rmd
Made graph for basis functions, need to change order of last paragraphs to fit the location of the plot

---
title: '**The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models**'
subtitle: _Beyond repeated measures ANOVA and Linear Mixed Models_
author:
- Ariel Mundo^[Department of Biomedical Engineering, University of Arkansas, Fayetteville]
- Timothy J. Muldoon^[Department of Biomedical Engineering, University of Arkansas,
  Fayetteville]
- John R. Tipton^[Department of Mathematical Sciences, University of Arkansas, Fayetteville]
output:
  bookdown::word_document2:
    fig_caption: yes
    keep_md: yes
  bookdown::pdf_document2:
    keep_tex: yes
    fig_caption: yes
    extra_dependencies: ["subfig","breqn"]
  bookdown::html_document2: default
  link-citations: yes
  css: style.css
  '': default
csl: Elsevier.csl
bibliography: refs.bib
---

@@

```{r,basis-functions}
library(mgcv)
library(mvnfast)
  library(nlme)
library(patchwork)
  library(ggsci)
library(tidyverse)

@@

#original bases
p1<-ggplot(basis_plot,aes(x=time,y=value_orig,colour=factor(Basis)))+
  geom_line()+geom_point()+
  facet_grid(~Basis)+
  theme_classic()+
  scale_color_aaas()

#basis multiplied by the coefficient
p2<-ggplot(basis_plot,aes(x=time,y=mod_val,colour=factor(Basis)))+
  geom_line()+geom_point()+
  facet_grid(~Basis)+
  theme_classic()+
  scale_color_aaas()

#heatmap of the coefficient
p3<-ggplot(basis_plot,aes(x=Basis,y=Basis,fill=cof))+
  geom_tile(aes(color='black'),size=2)+
  geom_tile(size=2)+
  scale_fill_gradient(low = "white", high = "red")+ 
  geom_text(aes(label=round(cof,1)))+
  theme_classic()

p4<-ggplot(data=dat,aes(x=time,y=y))+
  geom_point()+
  geom_line(data=spl,aes(x=time,y=spl))+
  scale_color_aaas()+
  theme_classic()

p1+p2+p3+p4+plot_annotation(tag_levels='A')
#plotting
#plot(dat$y~dat$time)

#matplot(dat$time, model_matrix[,-1], type = "l", lty = 2, add = T)

#lines(y_pred ~ dat$time, col = "red", lwd = 2)

>>>>

---
title: '**The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models**'
subtitle: _Beyond repeated measures ANOVA and Linear Mixed Models_
author:
- Ariel Mundo^[Department of Biomedical Engineering, University of Arkansas, Fayetteville]
- Timothy J. Muldoon^[Department of Biomedical Engineering, University of Arkansas,
  Fayetteville]
- John R. Tipton^[Department of Mathematical Sciences, University of Arkansas, Fayetteville]
output:
  bookdown::html_document2: default
  bookdown::pdf_document2:
    keep_tex: yes
    fig_caption: yes
    extra_dependencies: ["subfig","breqn"]
  bookdown::word_document2:
    fig_caption: yes
    keep_md: yes
  link-citations: yes
  css: style.css
  '': default
csl: Elsevier.csl
bibliography: refs.bib
---

@@

To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure @\ref(fig:quadratic-response). The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits the trend trough the timeline.  Because there are six timepoints for this group, the use of five equally spaced knots ensures that the model will not have computational issues. Asking the model to use five knots to construct the smooth term means that it will have _four_ effective basis functions (as one will be used for the intercept). The choice of basis functions is already optimized in the package _mgcv_ depending on the number of knots. In Panel A of Figure \@ref(fig:basis-functions-plot), the four basis functions are shown. Each of the basis functions is composed of six different points (because there are six points on the timeline). To control the wigliness of the fit, each of the basis functions of Panel A is penalized, or multipled by a coefficient according to the penalty matrix of Panel B. In other words, the six points of each basis will be multiplied by the corresponding coefficient, thereby increasing or decreasing the original basis functions of Panel B. In Figure \@ref(fig:basis-functions-plot), Panel C shows the resulting penalized basis functions. Note that the penalizetion for basis 1 has resulted in a decrease of its overall value (because the coefficient for that basis function is negative and less than 1); on the other hand, basis 3 has increased its value roughly 2.5 times. Finally, the penalized basis functions are added at each timepoint to produce the final value of the smooth term. The resulting smooth term for the effect of _time_ is shown in Panel D (orange line) along the simulated values per group, which appear as points.

```{r,basis-functions-plot,fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="**Basis functions for a single smoother for time with five knots. A: Basis functions for a single smoother for _time_ for the simulated data of Group 1 from Figure 2, the intercept basis is not shown. B: Penalty matrix for the basis functions. Each basis function is penalized by a coefficient which can be positive or negative. The coefficient determines the overall effect of each basis in the final smoother. C: Penalized basis functions. Each of the four basis functions of panel A has been penalized by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis. D. Smoother for _time_ and original datapoints. The smoother (line) is the result of the sum of each penalized basis function at each time point, simulated values for the group appear as points.**"}

library(mgcv)

@@

#original bases
sz<-1
p11<-ggplot(basis_plot,aes(x=time,y=value_orig,colour=as.factor(Basis)))+
  geom_line(size=sz,show.legend=FALSE)+
  geom_point(size=sz+1,show.legend = FALSE)+
  labs(y='Basis functions')+
  facet_wrap(~Basis)+
  theme_classic()+
  scale_color_aaas()

#basis multiplied by the coefficient
p12<-ggplot(basis_plot,aes(x=time,y=mod_val,colour=as.factor(Basis)))+
  geom_line(show.legend = FALSE,size=sz)+
  geom_point(show.legend = FALSE,size=sz+1)+
  labs(y='Penalized \n basis functions')+
  scale_y_continuous(breaks=seq(-1,1,1))+
  facet_wrap(~Basis)+
  theme_classic()+
  scale_color_aaas()

#heatmap of the coefficient
p13<-ggplot(basis_plot,aes(x=Basis,y=Basis,fill=cof))+
  geom_tile(aes(color='black'),size=sz+1,show.legend = FALSE)+
  geom_tile(size=sz+1)+
  scale_fill_gradient(low = "white", high = "red")+
  labs(x='Basis',y='Basis')+
  geom_text(aes(label=round(cof,2),size=10),show.legend = FALSE)+
  theme_classic()+
  theme(legend.title = element_blank())

#plotting simulated datapoints and spline  
p14<-ggplot(data=dat,aes(x=time,y=y))+
  geom_point(size=sz+1)+
  scale_color_aaas()+
  labs(y='Simulated \n response')+
  geom_line(data=spl,aes(x=time,y=spl),color="#B15731",size=sz+1)+
  theme_classic()

#Combining all
p11+p13+p12+p14+plot_annotation(tag_levels='A')&
  theme(plot.title = element_text(size = 30, 
                                face = "bold"),
     text=element_text(size=30))
